---
title: "HW 03"
author: "Nilotpal Rajbongshi"
format:
  html:
    embed-resources: true
toc: true
---execute:
  warning: false
  error: false
  message: false
---

## Setup

```{r setup}
if (!require("pacman")) 
  install.packages("pacman")

# use this line for installing/loading
pacman::p_load(tidyverse,
               ggplot2,
               scales,
               ggimage,
               ggtext,
               cowplot,
               forcats,
               stringr,
               patchwork)

devtools::install_github("tidyverse/dsbox")
```

## 1 - Du Bois challenge.

```{r}
#| label: label-me-1
income_data <- read_csv("data/income.csv") %>%
  rename(income_class = Class) %>%
  mutate(income_class = factor(income_class, levels = rev(unique(income_class))))

income_long <- income_data %>%
  pivot_longer(cols = c(Rent, Food, Clothes, Tax, Other),
               names_to = "category", values_to = "percent") %>%
filter(percent > 0.5) 

income_long$category <- factor(income_long$category,
                               levels = c("Other", "Tax", "Clothes", "Food", "Rent"))

category_colors <- c(
  Rent = "#161213",
  Food = "#a08294", 
  Clothes = "#d78879",   
  Tax = "#8e9aaf",      
  Other = "#ddead1"     
)

stacked_bar <- ggplot(income_long, aes(x = percent, y = income_class, fill = category)) +
  geom_col(color = "black", size = 0.3) +
  
  geom_text(
    aes(label = paste0(percent, "%"), color = category),
    position = position_stack(vjust = 0.5),
    size = 2.8,
    family = "Helvetica"
  ) +
  scale_color_manual(
    values = c("Rent" = "white", "Food" = "black", "Clothes" = "black", "Tax" = "black", "Other" = "black"),
    guide = "none"
  ) +

  scale_fill_manual(values = category_colors,
                    breaks = c("Rent", "Food", "Clothes", "Tax", "Other"),
                    labels = c("Rent", "Food", "Clothes", "Taxes", "Other Expenses & Savings"),
                    name = NULL) +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 100),
                     labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_family = "Helvetica") +
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "transparent", color = NA),
    plot.background = element_rect(fill = "transparent", color = NA),
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    axis.text.y = element_text(face = "bold", size = 10),
    legend.position = "top",
    legend.justification = "left"
  )

background_path <- "images/parchment.png"

final_plot <- ggdraw() +
  draw_image(background_path, scale = 3, x = 0, y = 0) +
  draw_plot(stacked_bar, x = 0, y = 0, width = 1, height = 1)

print(final_plot)

```

## 2 - COVID survey - interpret



## 3 - COVID survey - reconstruct
```{r, fig.width=15, fig.height=10}
#| label: COVID survey - reconstruct

# ─────────────────────────────────────────────────────────
# Load CSV  &  Clean raw data
# ─────────────────────────────────────────────────────────
covid_survey <- read_csv("data/covid-survey.csv", skip = 1)

# Confirm the shape
dim(covid_survey)

# drop rows where EVERYTHING except response_id is NA
covid_survey <- covid_survey %>%
  filter(!if_all(-response_id, is.na))

# Confirm the shape again
dim(covid_survey)

# recode explanatory variables
covid_survey <- covid_survey %>%
  mutate(
    exp_already_vax = recode(exp_already_vax, `0` = "No", `1` = "Yes"),
    exp_flu_vax = recode(exp_flu_vax, `0` = "No", `1` = "Yes"),
    exp_profession = recode(exp_profession, `0` = "Medical", `1` = "Nursing"),
    exp_gender = recode(exp_gender, `0` = "Male", `1` = "Female", `3` = "Non-binary third gender", `4` = "Prefer not to say"),
    exp_race = recode(exp_race, `1` = "American Indian / Alaskan Native", `2` = "Asian", `3` = "Black / African American", `4` = "Native Hawaiian / Other Pacific Islander", `5` = "White"),
    exp_ethnicity = recode(exp_ethnicity, `1` = "Hispanic / Latino", `2` = "Non-Hispanic/Non-Latino"),
    exp_age_bin = recode(exp_age_bin, `0` = "<20", `20` = "21-25", `25` = "26-30", `30` = ">30")
  )

dim(covid_survey)

# ─────────────────────────────────────────────────────────
# Pivot to long & summarise (mean, 10th, 90th)
# ─────────────────────────────────────────────────────────
covid_survey_longer <- covid_survey %>%
  pivot_longer(
    cols = starts_with("exp_"),
    names_to = "explanatory",
    values_to = "explanatory_value"
  ) %>%
  filter(!is.na(explanatory_value)) %>%
  pivot_longer(
    cols = starts_with("resp_"),
    names_to = "response",
    values_to = "response_value"
  )

covid_survey_longer

covid_survey_summary_stats_by_group <- covid_survey_longer %>%
  group_by(explanatory, explanatory_value, response) %>%
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.10, na.rm = TRUE),
    high = quantile(response_value, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

covid_survey_summary_stats_by_group

covid_survey_summary_stats_all <- covid_survey_longer %>%
  group_by(response) %>%
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.10, na.rm = TRUE),
    high = quantile(response_value, 0.90, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    explanatory = "All",
    explanatory_value = ""
  )

covid_survey_summary_stats_all

covid_survey_summary_stats <- bind_rows(
  covid_survey_summary_stats_all,
  covid_survey_summary_stats_by_group
)

covid_survey_summary_stats

# ─────────────────────────────────────────────────────────
# Facet-strip label vectors
# ─────────────────────────────────────────────────────────
response_labels <- c(
  resp_safety = "Based on my understanding,\nI believe the vaccine is safe",
  resp_confidence_science = "I am confident in the\nscientific vetting process\nfor the new COVID vaccines",
  resp_feel_safe_at_work = "Getting the vaccine\nwill make me feel safer at work",
  resp_will_recommend = "I will recommend the vaccine\nto family, friends,\nand community members",
  resp_trust_info = "I trust the information that\nI have received about the vaccines",
  resp_concern_safety = "I am concerned about the\nsafety and side effects\nof the vaccine"
)

explanatory_labels <- c(
  exp_age_bin = "Age",
  exp_gender = "Gender",
  exp_race = "Race",
  exp_ethnicity = "Ethnicity",
  exp_profession = "Profession",
  exp_already_vax = "Had COVID vaccine",
  exp_flu_vax = "Had flu vaccine this year",
  All = "All"
)

options(repr.plot.width  = 18,
        repr.plot.height = 15)

covid_survey_summary_stats <- covid_survey_summary_stats %>% 
  group_by(explanatory) %>% 
  mutate(explanatory_value = fct_rev(fct_inorder(explanatory_value))) %>% 
  ungroup()

resp_lab_wrapped <- setNames(
  str_wrap(unname(response_labels), width = 25),
  names(response_labels)
)

# ─────────────────────────────────────────────────────────
# Final faceted plot
# ─────────────────────────────────────────────────────────
ggplot(covid_survey_summary_stats,
       aes(x = mean, y = explanatory_value)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = low, xmax = high), height = .15) +
  facet_grid(
    explanatory ~ response,
    scales  = "free_y",
    space   = "free_y",
    labeller = labeller(
      response    = resp_lab_wrapped,
      explanatory = explanatory_labels
    )
  ) +                         
  scale_x_continuous(breaks = 1:5) +
  coord_cartesian(xlim = c(1, 5)) +
  theme_minimal(base_size = 9) +
  theme(
    strip.background   = element_rect(fill = "gray90", colour = "black"),
    strip.text.y       = element_text(angle = 0, hjust = 0),
    axis.title.y       = element_blank(),
    axis.title.x = element_text(colour = "black"), 
    plot.margin        = margin(12, 12, 12, 12),
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank()
  ) +
  labs(
    x = "Mean Likert score\n(Error bars show 10th–90th percentiles)"
  )

```
## 4 - COVID survey - re-reconstruct
```{r, fig.width=15, fig.height=10}

#| label: COVID survey - reconstruct

# ─────────────────────────────────────────────────────────
# Load CSV  &  Clean raw data
# ─────────────────────────────────────────────────────────
covid_survey <- read_csv("data/covid-survey.csv", skip = 1)

# Confirm the shape
dim(covid_survey)

# drop rows where EVERYTHING except response_id is NA
covid_survey <- covid_survey %>%
  filter(!if_all(-response_id, is.na))

# Confirm the shape again
dim(covid_survey)

# recode explanatory variables
covid_survey <- covid_survey %>%
  mutate(
    exp_already_vax = recode(exp_already_vax, `0` = "No", `1` = "Yes"),
    exp_flu_vax = recode(exp_flu_vax, `0` = "No", `1` = "Yes"),
    exp_profession = recode(exp_profession, `0` = "Medical", `1` = "Nursing"),
    exp_gender = recode(exp_gender, `0` = "Male", `1` = "Female", `3` = "Non-binary third gender", `4` = "Prefer not to say"),
    exp_race = recode(exp_race, `1` = "American Indian / Alaskan Native", `2` = "Asian", `3` = "Black / African American", `4` = "Native Hawaiian / Other Pacific Islander", `5` = "White"),
    exp_ethnicity = recode(exp_ethnicity, `1` = "Hispanic / Latino", `2` = "Non-Hispanic/Non-Latino"),
    exp_age_bin = recode(exp_age_bin, `0` = "<20", `20` = "21-25", `25` = "26-30", `30` = ">30")
  )

dim(covid_survey)

# ─────────────────────────────────────────────────────────
# Pivot to long & summarise (mean, 10th, 90th)
# ─────────────────────────────────────────────────────────
covid_survey_longer <- covid_survey %>%
  pivot_longer(
    cols = starts_with("exp_"),
    names_to = "explanatory",
    values_to = "explanatory_value"
  ) %>%
  filter(!is.na(explanatory_value)) %>%
  pivot_longer(
    cols = starts_with("resp_"),
    names_to = "response",
    values_to = "response_value"
  )

covid_survey_longer

# Changes according to 
covid_survey_summary_stats_by_group <- covid_survey_longer %>%
  group_by(explanatory, explanatory_value, response) %>%
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.25, na.rm = TRUE),
    high = quantile(response_value, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

covid_survey_summary_stats_by_group

covid_survey_summary_stats_all <- covid_survey_longer %>%
  group_by(response) %>%
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.25, na.rm = TRUE),
    high = quantile(response_value, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    explanatory = "All",
    explanatory_value = ""
  )

covid_survey_summary_stats_all

covid_survey_summary_stats <- bind_rows(
  covid_survey_summary_stats_all,
  covid_survey_summary_stats_by_group
)

covid_survey_summary_stats

# ─────────────────────────────────────────────────────────
# Facet-strip label vectors
# ─────────────────────────────────────────────────────────
response_labels <- c(
  resp_safety = "Based on my understanding,\nI believe the vaccine is safe",
  resp_confidence_science = "I am confident in the\nscientific vetting process\nfor the new COVID vaccines",
  resp_feel_safe_at_work = "Getting the vaccine\nwill make me feel safer at work",
  resp_will_recommend = "I will recommend the vaccine\nto family, friends,\nand community members",
  resp_trust_info = "I trust the information that\nI have received about the vaccines",
  resp_concern_safety = "I am concerned about the\nsafety and side effects\nof the vaccine"
)

explanatory_labels <- c(
  exp_age_bin = "Age",
  exp_gender = "Gender",
  exp_race = "Race",
  exp_ethnicity = "Ethnicity",
  exp_profession = "Profession",
  exp_already_vax = "Had COVID vaccine",
  exp_flu_vax = "Had flu vaccine this year",
  All = "All"
)

options(repr.plot.width  = 18,
        repr.plot.height = 15)

covid_survey_summary_stats <- covid_survey_summary_stats %>% 
  group_by(explanatory) %>% 
  mutate(explanatory_value = fct_rev(fct_inorder(explanatory_value))) %>% 
  ungroup()

resp_lab_wrapped <- setNames(
  str_wrap(unname(response_labels), width = 25),
  names(response_labels)
)

# ─────────────────────────────────────────────────────────
# Final faceted plot
# ─────────────────────────────────────────────────────────
ggplot(covid_survey_summary_stats,
       aes(x = mean, y = explanatory_value)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = low, xmax = high), height = .15) +
  facet_grid(
    explanatory ~ response,
    scales  = "free_y",
    space   = "free_y",
    labeller = labeller(
      response    = resp_lab_wrapped,
      explanatory = explanatory_labels
    )
  ) +                         
  scale_x_continuous(breaks = 1:5) +
  coord_cartesian(xlim = c(1, 5)) +
  theme_minimal(base_size = 9) +
  theme(
    strip.background   = element_rect(fill = "gray90", colour = "black"),
    strip.text.y       = element_text(angle = 0, hjust = 0),
    axis.title.y       = element_blank(),
    axis.title.x = element_text(colour = "black"), 
    plot.margin        = margin(12, 12, 12, 12),
    panel.grid.major   = element_blank(),
    panel.grid.minor   = element_blank()
  ) +
  labs(
    x = "Mean Likert score\n(Error bars show 25th–75th percentiles)"
  )
```
Using the 25th–75th percentile range instead of the 10th–90th makes the plot more focused on the core responses, reducing the influence of outlines. This results in shorter error bars, making differences between groups appear more distinct and easier to compare. While the overall trends and conclusions remain consistent with the 10th–90th version, the revised plot provides a clearer view of the central tendency and typical variability in responses. It helps emphasize where most respondents fall, rather than highlighting the full range of opinions.

## 5 - COVID survey - another view
```{r}
likert_levels <- c("Strongly Agree", "Somewhat Agree", "Neutral", "Somewhat Disagree", "Strongly Disagree")
likert_colors <- c(
  "Strongly Agree" = "#006400",
  "Somewhat Agree" = "#66CD00",
  "Neutral" = "#FFFF66",
  "Somewhat Disagree" = "#FFA500",
  "Strongly Disagree" = "#B22222"
)

# Define descriptive question labels
response_labels <- c(
  resp_safety = "Based on my understanding,\nI believe the vaccine is safe",
  resp_confidence_science = "I am confident in the\nscientific vetting process\nfor the new COVID vaccines",
  resp_feel_safe_at_work = "Getting the vaccine\nwill make me feel safer at work",
  resp_will_recommend = "I will recommend the vaccine\nto family, friends,\nand community members",
  resp_trust_info = "I trust the information that\nI have received about the vaccines",
  resp_concern_safety = "I am concerned about the\nsafety and side effects\nof the vaccine"
)

# Prepare and summarize data
likert_summary <- covid_survey %>%
  pivot_longer(cols = starts_with("resp_"), names_to = "question", values_to = "response_value") %>%
  filter(!is.na(response_value)) %>%
  mutate(
    response_label = factor(response_value, levels = 1:5, labels = likert_levels),
    question = recode(question, !!!response_labels),
    question = factor(question, levels = response_labels)
  ) %>%
  count(question, response_label, name = "n") %>%
  group_by(question) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Create diverging version of the data
likert_diverging <- likert_summary %>%
  mutate(pct = if_else(response_label %in% likert_levels[1:2], -pct, pct))

# Diverging bar plot
diverging_plot <- ggplot(likert_diverging, aes(x = pct, y = fct_rev(question), fill = response_label)) +
  geom_col(width = 0.7) +
  scale_x_continuous(
  breaks = seq(-0.75, 0.75, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
  ) +
  scale_fill_manual(values = likert_colors) +
  labs(x = "Proportion of Responses", y = NULL) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 8),
    legend.title = element_blank(),
    legend.key.width = unit(0.3, "cm"),
    legend.key.height = unit(0.3, "cm"),
    legend.spacing.x = unit(0.3, "cm"),
    plot.margin = margin(5, 5, 5, 5)
  )

# 100% stacked bar plot
stacked_plot <- ggplot(likert_summary, aes(x = fct_rev(question), y = pct, fill = response_label)) +
  geom_col(position = "fill", width = 0.7) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = likert_colors) +
  labs(x = NULL, y = "Proportion of Responses") +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none",
    plot.margin = margin(5, 5, 5, 5)
  )

# Combine plots with shared legend
(diverging_plot + stacked_plot) +
  plot_layout(ncol = 2, widths = c(1.3, 1.3), guides = "collect") +
  plot_annotation(
    title = "What people had to say about COVID-19 Vaccines",
    subtitle = "Interpreting the survey using a Diverging bar plot and a 100% bar plot",
    theme = theme(plot.title = element_text(face = "bold", size = 16),
                  plot.subtitle = element_text(size = 12),
                  legend.position = "top")
  )
```
